Navigating Hostility: The Effect of Nonprofit Transparency and Accountability on Donor Preferences in the Face of Shrinking Civic Space

### 2025-01-22 manuscript/manuscript.qmd
MEGA EDITS for word count issues

---
title: "Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy"
short-title: "Why Donors Donate"
date: "2025-01-22"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Utah State University
  url: https://occasionaldivergences.com/
  email: marc.dotson@usu.edu
  orcid: 0000-0001-7846-3660
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Governments across the world have increasingly used laws to restrict the work of nonprofits, which has led to a reduction in public or official foreign aid directed towards these groups. Many international nonprofits, in response, have turned to individual donors to offset the loss of traditional funding. What are individual donors' preferences regarding donating to legally besieged nonprofits abroad? We conducted a conjoint experiment on a nationally representative sample of likely donors in the US and found that learning about host government criticism and legal restrictions on nonprofits decreases individuals' preferences to donate to them. However, organizational features such as financial transparency and accountability can protect against this dampening effect. Our results have important implications both for understanding private international philanthropy and how nonprofits can better frame their fundraising appeals at a time when they are facing restrictive civic spaces and hostile governments abroad.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/andrewheiss/silent-skywalk>
keywords:
  - philanthropy
  - conjoint experiments
  - donor heuristics
  - repression
  - NGOs
  - civil society
  - nonprofits
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "9,300"
---

@@

Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Over the last two decades, however, NGOs face restrictive civic spaces abroad [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have increasingly used legal measures to create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. For instance, in 2012 Russia passed its infamous "Foreign Agent" law (Federal Law No. 121-F2), which forced organizations receiving foreign funding and engaging in political activities to register as foreign agents, which forced hundreds of organizations to close down [@Kirova:2024]. These developments are not limited to authoritarian regimes. Hybrid and democratic states have also restricted NGOs' operations within their territories. Between 2014 and 2019, the Indian government banned 14,500 NGOs from receiving money from abroad, including prominent international nonprofits such as the Ford Foundation and Greenpeace [@PressTrustofIndia:2019]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. This phenomenon of closing civic space has a large deterrent effect on both advocacy *and* service-delivery NGOs [@SpringmanMaleskyRight:2022].

@@

The amount of private foreign aid is not insignificant---from 2000--2010, private giving by US entities towards international causes doubled in real terms [@DesaiKharas:2018, 505]. In 2018, this amounted to \$22.88 billion in individual donations earmarked specifically for international affairs INGOs [@GivingUSA:2019]. This turn to private donors opens INGOs up to new dynamics in seeking out funds. Individual donor behavior is unlike official donor behavior [@DesaiKharas:2018]. They have different preferences---while official donors channel funds to countries with better institutional quality or to reward governmental performance [@Bermeo:2011; @DietrichWright:2015], individual donors are more inclined to donate to NGOs in countries experiencing a humanitarian crisis [@DesaiKharas:2018]. As such, individual donors may have separate metrics of effectiveness or impact that can differ from the benchmarks that large donor agencies rely on, and INGOs may need to adapt their fundraising strategies to a different set of potential donors.

@@

Existing research shows that organizational accountability [@Becker:2018; @Tremblay-BoirePrakash:2017]  and financial transparency [@HarrisNeely:2021; @SaxtonKuoHo:2012; @Vaccaro:2006; @VaccaroMadsen:2006] play an important role in motivating individual donors to donate to NGOs. We extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with research on the consequences of closing civic space on NGOs and donors [@DupuyPrakash:2018; @RightSpringmanWibbels:2022; @FransenDupuyHinfelaar:2021; @SundstromHenrySperling:2022; @Herrold:2020; @ChaudhryHeiss:2021]. We argue that organization characteristics---such as an NGO's transparency and accountability practices---can serve as important heuristics that influence foreign private donors' preferences even during closing civic space.

@@

Our findings make two main contributions. First, existing research on privately-provided foreign aid primarily looks at organizations like foundations and large donor NGOs, and not private individuals [@ButheMajordeMelloeSouza:2012; @Koch:2009; @DesaiKharas:2018], and studies that do examine individual giving behavior typically only examine wealthy individuals [@USTrust2014]. Our research expands work on private international philanthropy by exploring how non-elite individuals decide to donate to NGOs focused on international affairs. Second, research shows that NGO leaders often have difficulty in attracting new donors and also find it challenging to motivate current donors [@GaudreauCao:2015]. Examining organization-level factors that make the average donor increase their charitable giving, especially when facing restricted civic space abroad, can help NGOs better frame their funding appeals. Moreover, our results provide guidance to organizations facing increased antagonism by governments abroad and point to possible methods of adapting to worsening or shrinking civic space.

@@

A large body of work in philanthropy explores the factors that shape private donor motivation, including altruism, reputational benefits, and alignment with personal values [@BilodeauSlivinski:1997; @Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to *domestic* organizations instead of international causes [@BekkersWiepking:2011; @Tremblay-BoirePrakash:2017]. Giving to local organizations involves more observable results, while the benefits of donating to organizations working abroad are less visible and more removed from donors [@CasaleBaumann:2015]. @Tremblay-BoirePrakash:2017 confirm this, finding that "donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad" [-@Tremblay-BoirePrakash:2017, 644].

@@

But as INGOs became increasingly powerful actors in global policy circles, they faced a number of criticisms. Many organizations struggle with the conflicting imperatives of being large organizations that can operate efficiently at scale, while maintaining the grassroots contacts and consensus building that contributed to their early successes [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008; @Bano:2012; @Rahman:2006]. As a result, many nonprofits sought to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organization-level factors that can potentially impact nonprofit efficacy, including organizational and governance structure, leadership, strategy, transparency, and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries and devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy involves whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning, and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organization-level factors) are often not visible to the average individual (non-elite) donor, especially in the context of international NGOs. Therefore, in this paper, we focus on the two organization-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, these donors rely on various easily available cues and heuristics when deciding to support an NGO. Compared to other organization-level factors, news about financial mismanagement and lack of transparency and accountability may be more easily available. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the Oil-for-Food program, initiated by the United Nations in 1996 to allow Iraq to sell enough oil to pay for food and other necessities for its population, which had been suffering from sanctions imposed on Iraq since the First Gulf War. However, in 2004, the program's widespread financial fraud, limited distribution of food by nonprofits, and lack of accountability was widely publicized. Numerous other such cases---covered in @GibelmanGelman:2001 and @GibelmanGelman:2004---show how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust in and donations to these nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watch dogs such as Charity Navigator and Better Business Bureau Wise Giving Alliance often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways, such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@Vaccaro:2006; @VaccaroMadsen:2006].

@@

Accountability is another factor that can impact donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also encompasses how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders, or "any group or individual who can affect or is affected by the achievement of an organization's objectives," [@MitchellAgleWood:1997, 854] want to be assured that their donations or support or support of an organization will make a real difference. More effective accountability can encourage NGOs to become more closely aligned with community practices, provide greater assurances to stakeholders, and help expand their support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgements or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, deteriorated reputation, and lower perceived quality [@Becker:2018].

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this paper, such as the Russian Foreign Agent law, are not grounded in international legal principles and do not give nonprofits the freedom of association or access to resources [@Voule:2023; @Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

@@

Some NGOs have adapted to these legal crackdowns by recreating their organizations and changing their mission to avoid directly confronting the government. Analyzing the impact of anti-NGO laws in Bangladesh and Zambia, scholars have found that when faced with government restrictions, NGOs with broad missions restricted their strategies to service work and abandoned their advocacy work. Those that continued to engage in transnational advocacy changed the targets of their advocacy, the issues they focused on, and the language they used [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer groups may end up focusing on contentious causes, instead focusing on issues that are less likely to invoke the ire of the host government, including in areas such as health and education.

The changes in civic environments impacting NGO activity have already influenced official donor responses. Scholars have found that while multilateral donors do not reduce aid in response to anti-NGO laws, these laws are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022].

@@

It is possible to study the effects of both organizational and structural heuristics on donor decision making individually, but disentangling the interaction of the two heuristics from individual donor characteristics adds complex dimensionality and makes more standard experimental work difficult, costly, and statistically fraught. To address this, we use conjoint analysis to simultaneously test individual donor preferences in the face of both organizational and structural heuristics. Conjoint analysis is generally applicable whenever individuals face complex, multi-dimensional tradeoffs that make it difficult (or even impossible) to explicitly elicit preferences. Conjoint designs have been used extensively in political science to study complex preferences for elections, laws, public policy, and public opinion [@HainmuellerHopkinsYamamoto:2014; @Mummolo:2016; @KertzerRenshonYarhi-Milo:2021; @BansakHainmuellerHopkins:2021; @BremerBurgisser:2023], and conjoint experiments are becoming more common in nonprofit studies, particularly for studying individual donor behavior [@SilviaChildWitesman:2023; @HirschmannMoritzBlock:2022; @BachkeAlfnesWik:2014; @Kottasz:2004]. @KnudsenJohannesson:2018 argue that research related to respondent opinions requires conjoint analysis because it accounts for the masking effects of factors as well as the complexity of a respondent's judgment of credibility. Because we also want to capture the complexity of a respondent's judgment of philanthropic preferences, we use a conjoint experiment to measure the causal effect of both organizational and structural heuristics on individuals' preferences for donating to international NGOs.

A conjoint research design also allows us to calculate treatment effects even if respondents do not see every combination of organization attributes, and correspondingly provides a substantial increase in statistical power. In standard factorial experimental designs, participants would ordinarily need to be shown all possible combinations of experimental attributes. In our experiment, we presented respondents with six possible treatments with randomized attributes for each treatment, yielding 576 possible unique combinations of features. Even with respondents answering twelve iterations of the experiment, not every combination was seen. However, as long as all the possible organizational attributes are well randomized and there are no systematic biases toward specific options (i.e., more respondents select the first option because it is the first) or toward earlier iterations of attribute choices (i.e., respondents are more careful and attentive for the first hypothetical organization than the last), we can pool all observations together for specific attributes of interest while marginalizing across all other attributes [@KertzerRenshonYarhi-Milo:2021]. This allows us to (1) estimate the effect of each experimental treatment even if some unique combinations of attributes were unseen, and (2) use a much smaller sample size than would be required in a more traditional factorial design.

@@

[^prereg]: For narrative purposes, we do not explore all our preregistered hypotheses in this study---many of them are nested inside each other (e.g., the effect of branding and issue area and transparency and crackdown simultaneously). We instead focus here on the simplest, non-nested hypotheses. Following @WillrothAtherton:2024, we outline all the deviations from our preregistration protocol in the appendix.

@@

We fielded our survey experiment through Centiment, which recruits representative samples of paid (and highly engaged) survey participants online. To see how varying NGO characteristics influence the decision to donate, our sample was representative of the population of people who are likely to donate to charity. We asked potential participants a screening question about their philanthropic behavior early in the survey---if a participant responded that they give once every few years or never, they were disqualified from the study. We also included an attention check question early in the survey and removed respondents who failed the question. Importantly, these screening questions were presented *prior to* the experimental manipulation to avoid post-treatment bias [@MontgomeryNyhanTorres:2018a]. Following screening, we received 1,016 viable responses. Tables A1, A2, and A3 provide a detailed breakdown of the individual characteristics of our sample. In general, respondents were well balanced across all pre-treatment characteristics, including gender, age, education, income, and attitudes toward charity. Additionally, Table A4 compares sample characteristics with nationally representative estimates from the 2019 Current Population Survey (CPS). Our sample matches national proportions of age, marital status, and education, and contains slightly more men and slightly higher income levels than the general population. Due to the initial screening question that limited participants to those with an interest in charitable activity, respondents are far more likely to have donated, volunteered, or voted in the past year than population CPS estimates. Our reported causal effects are therefore not representative of the entire population and instead are only valid for people interested in charitable giving.

@@

[^real-names]: We used real organization names rather than hypothetical, fictionalized names to allow the experimental task to better mirror reality [@McDonald:2020] and to reduce respondent cognitive burden when reflecting on organization characteristics [@MajnemerMeibauer:2022].

After presenting respondents with a set of three randomly generated organizations, we asked them which of the three they would be willing to donate to, along with an option for no selection (see @tbl-example for an example).This choice indicated donation intention and reflected the option the respondent preferred most within that set. We then repeated the process eleven more times for a total of twelve randomized iterations of hypothetical combinations of attributes, resulting in 12,192 completed experimental tasks (12 iterations × 1,016 respondents). This iterative process across twelve experimental tasks is typical of conjoint analysis and allows us to estimate individual-level differences using a multilevel model and determine both aggregate respondent preferences and the causal effect of each experimental feature on those preferences.

@@

We present the posterior distributions of the marginal means and AMCEs for each of our experimental conditions in @fig-all-results and provide posterior medians, credible intervals, and other model diagnostics in Tables A6 and A8. Because AMCEs are relative statements (i.e., contrasts between one feature level and a reference level), we try to use logical reference levels: for binary treatments like transparency and accountability, we calculate the difference between false and true levels; for ordered treatments like host country relationship, we calculate the differences between different levels of crackdown compared to no crackdown. To avoid imposing an artificial order on other unordered treatment variables, we report both marginal means and AMCEs [@LeeperHoboltTilley:2020].

>>>>

---
title: "Navigating Hostility: The Effect of Nonprofit Transparency and Accountability on Donor Preferences in the Face of Shrinking Civic Space"
short-title: "Navigating Hostility"
date: "2025-01-22"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  url: http://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
  orcid: 0000-0003-3506-7590
- name: Marc Dotson
  affiliation: Utah State University
  url: https://occasionaldivergences.com/
  email: marc.dotson@usu.edu
  orcid: 0000-0001-7846-3660
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  url: https://www.andrewheiss.com/
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Governments across the world have increasingly used laws to restrict the work of nonprofits, which has led to a reduction in public or official foreign aid directed towards these groups. Many international nonprofits, in response, have turned to individual donors to offset the loss of traditional funding. What are individual donors' preferences regarding donating to legally besieged nonprofits abroad? We conducted a conjoint experiment on a nationally representative sample of likely donors in the US and found that learning about host government criticism and legal restrictions on nonprofits decreases individuals' preferences to donate to them. However, organizational features such as financial transparency and accountability can protect against this dampening effect. Our results have important implications both for understanding private international philanthropy and how nonprofits can better frame their fundraising appeals at a time when they are facing restrictive civic spaces and hostile governments abroad.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/andrewheiss/silent-skywalk>
keywords:
  - philanthropy
  - conjoint experiments
  - donor heuristics
  - repression
  - NGOs
  - civil society
  - nonprofits
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/silent-skywalk>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
word-count: "10,149"  # Subtract 77 for the final statements at the end
---

@@

Non-governmental organizations (NGOs) are a core component of global governance and service delivery. Over the last two decades, however, NGOs face restrictive civic spaces abroad [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have increasingly used legal measures to create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. For instance, in 2012 Russia passed its infamous "Foreign Agent" law (Federal Law No. 121-F2), which forced organizations receiving foreign funding and engaging in political activities to register as foreign agents, which forced hundreds of organizations to close down [@Kirova:2024]. These developments are not limited to authoritarian regimes. Hybrid and democratic states have also restricted NGOs' operations within their territories. Between 2014 and 2019, the Indian government banned 14,500 NGOs from receiving money from abroad, including prominent international nonprofits such as the Ford Foundation and Greenpeace [@PressTrustofIndia:2019]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. This phenomenon of closing civic space has a large deterrent effect on both advocacy and service-delivery NGOs [@SpringmanMaleskyRight:2022].

@@

The amount of private foreign aid is not insignificant---from 2000--2010, private giving by US entities towards international causes doubled in real terms [@DesaiKharas:2018, 505]. This turn to private donors opens INGOs up to new dynamics in seeking out funds. Individual donor behavior is unlike official donor behavior [@DesaiKharas:2018]. They have different preferences---while official donors channel funds to countries with better institutional quality or to reward governmental performance [@Bermeo:2011; @DietrichWright:2015], individual donors are more inclined to donate to NGOs in countries experiencing a humanitarian crisis [@DesaiKharas:2018]. As such, individual donors may have separate metrics of effectiveness or impact that can differ from the benchmarks that large donor agencies rely on, and INGOs may need to adapt their fundraising strategies to a different set of potential donors.

@@

Existing research shows that organizational accountability [@Becker:2018; @Tremblay-BoirePrakash:2017]  and financial transparency [@HarrisNeely:2021; @SaxtonKuoHo:2012; @VaccaroMadsen:2006] play an important role in motivating individual donors to donate to NGOs. We extend this literature by evaluating how these organization-level characteristics influence private donor behavior in an era of shrinking civic space. In considering this question, we engage with research on the consequences of closing civic space on NGOs and donors [@DupuyPrakash:2018; @RightSpringmanWibbels:2022; @FransenDupuyHinfelaar:2021; @SundstromHenrySperling:2022; @Herrold:2020; @ChaudhryHeiss:2021]. We argue that organization characteristics---such as an NGO's transparency and accountability practices---can serve as important heuristics that influence foreign private donors' preferences even during closing civic space.

@@

Our findings make two main contributions. First, existing research on privately-provided foreign aid primarily looks at organizations like foundations and large donor NGOs, and not private individuals [@ButheMajordeMelloeSouza:2012; @DesaiKharas:2018], and studies that do examine individual giving behavior typically only examine wealthy individuals [@USTrust2014]. Our research expands work on private international philanthropy by exploring how non-elite individuals decide to donate to NGOs focused on international affairs. Second, research shows that NGO leaders often have difficulty in attracting new donors and also find it challenging to motivate current donors [@GaudreauCao:2015]. Examining organization-level factors that make the average donor increase their charitable giving, especially when facing restricted civic space abroad, can help NGOs better frame their funding appeals. Moreover, our results provide guidance to organizations facing increased antagonism by governments abroad and point to possible methods of adapting to worsening or shrinking civic space.

@@

A large body of work in philanthropy explores the factors that shape private donor motivation, including altruism, reputational benefits, and alignment with personal values [@Wiepking:2010; @BekkersWiepking:2011]. However, this research has overwhelmingly looked at individual donor behavior and motivation for giving to *domestic* organizations instead of international causes [@BekkersWiepking:2011; @Tremblay-BoirePrakash:2017]. Giving to local organizations involves more observable results, while the benefits of donating to organizations working abroad are less visible and more removed from donors [@CasaleBaumann:2015]. @Tremblay-BoirePrakash:2017 confirm this, finding that "donors are more likely to donate to a charity operating locally than to a charity providing identical service abroad" [-@Tremblay-BoirePrakash:2017, 644].

@@

But as INGOs gained more power in global policy circles, they faced a number of criticisms. Many organizations struggle to balance efficient large-scale operations with the grassroots connections and consensus-building that drove their early success [@MitchellSchmitzBruno-vanVijfeijken:2020; @Jalali:2008]. As a result, many INGOs worked to establish and enhance their accountability, responsiveness, and legitimacy [@edwardshulme; @GourevitchLake:2012]. These actions were a result of many organization-level factors that can potentially impact nonprofit efficacy, including organizational and governance structure, leadership, strategy, transparency, and accountability. Organizational and governance structures define how decisions are made and who participates in decision making. While some nonprofits follow more democratic governance models---such as expanded membership to Global South countries and devolving more decision-making powers to the regional or country levels, others have a more centralized structure [@Wong:2012; @MitchellSchmitzBruno-vanVijfeijken:2020, 155]. Leadership can be defined as how leaders "handle the external and internal politics of making their organizations more effective and responsive," and what sort of organizational cultures they create [@MitchellSchmitzBruno-vanVijfeijken:2020, 178]. Finally, strategy involves whether the organization has a long-term strategic plan that emphasizes sustainability, innovation, learning, and adaptation [@MitchellSchmitzBruno-vanVijfeijken:2020, 100].

However, organizational structure, leadership, and strategy (among other organization-level factors) are often not visible to the average individual donor. We therefore focus on two organization-level factors that are most visible to the average donor---financial transparency and accountability. Assessing an NGO's deservingness and efficiency demands substantial time and resources, which individual donors typically lack [@CrosonShang:2011; @Tremblay-BoirePrakash:2017]. As such, donors rely on cues and heuristics when deciding to support an NGO. Compared to other organization-level factors, news about financial mismanagement and lack of transparency and accountability may be more easily available. Many nonprofit scandals involve some form of financial wrongdoing [@PrakashGugerty:2010b]. Notable examples include the United Nations' Oil-for-Food program, aimed at alleviated Iraq's sanctions-induced hardships in the 1990s, which was marred by revelations of widespred financial fraud, limited food distribution, and lack of accountability in 2004. Numerous other cases [see @GibelmanGelman:2001 and @GibelmanGelman:2004] demonstrate how financial mismanagement, embezzlement, fraud, and lack of accountability can decrease public trust and reduce donations to nonprofits. Globally, around 5% of revenue is lost annually due to fraud [@ACFE:2021]. Misappropriation of assets and embezzlement schemes are among the most common forms of financial fraud [@LamotheTer-MkrtchyanRuddle:2023]. To save time and present easily accessible information, watchdogs such as Charity Navigator often rate charities on financial and accountability metrics and present this information in a quick, user-friendly format.

@@

We first examine the impact of two organizational practices---financial transparency and accountability---on individual donor preferences. Financial transparency, or the "degree of completeness of information provided by [organizations] to the [public] concerning [their] activities," [@VaccaroMadsen:2006, p. 146] is one tangible heuristic that donors turn to when making decisions. Transparency can involve proactive sharing of information with stakeholders, participation in data transparency platforms, avoiding financial misconduct, and exhibiting trustworthiness. NGOs can engage in financial transparency in a variety of ways, such as distribution of audited financial statements and voluntarily sharing information with third party intermediaries. Research shows that transparency is associated with greater donations, lower debt, and better governance [@HarrisNeely:2021; @HarrisPetrovitsYetman:2023; @SaxtonKuoHo:2012]. It can also enhance donor trust and confidence in organizations [@VaccaroMadsen:2006].

@@

Accountability also influences donor preferences. Accountability is the willingness of the organization to explain its action to shareholders [@CharityNavigator:2020] and build and manage relationships with stakeholders or client populations and donor groups. It also reflects how well an organization's mission is aligned with its resources [@Sloan:2009]. As NGOs gain more prominence, stakeholders seek assurance that their donations will make a real difference---effective accountability can encourage NGOs to become more closely aligned with community practices and help expand stakeholder support [@Murtaza:2012]. Problems with accountability usually occur when organizations ignore their stakeholders' preferences and use judgements or actions which may not align with one or more of their constituent groups [@McCambridge:2019]. The absence of accountability is associated with decreased public trust, deteriorated reputation, and lower perceived quality [@Becker:2018].

@@

While NGOs have control over organizational factors such as transparency and accountability and can work to improve these characteristics and public perception of them, NGOs may have little to no control over structural factors that can also influence philanthropy. These structural factors include political contexts in NGO host countries, especially a restrictive environment for civil society organizations. A global cascade of anti-NGO laws in recent decades has created barriers to entry, funding, and advocacy for civic organizations [@GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020; @Chaudhry:2022; @Heiss:2017; @DupuyRonPrakash:2016]. The use of such anti-NGO laws, also referred to as administrative crackdown [@Chaudhry:2022], is different from regulations that simply set standards for appropriate organizational behavior and set penalties for violations [@North:1990].[^violent-crackdown] However, the anti-NGO laws discussed in this paper, such as the Russian Foreign Agent law, are not grounded in international legal principles and do not give nonprofits the freedom of association or access to resources [@Robinson:2024]. As a result, NGOs must spend more time, effort, and resources to ensure their survival---at the expense of pursuing their missions.

@@

Some NGOs have adapted to these legal crackdowns by recreating their organizations and changing their mission to avoid directly confronting the government. Research on anti-NGO laws in Bangladesh and Zambia shows that NGOs with broad missions shifted from advocacy to service work, while those focused mainly on transnational advocacy altered their targets, issues, and language [@FransenDupuyHinfelaar:2021]. Similarly, in Russia, many NGOs working with foreign partners and lobbying the central government no longer consider it an effective strategy [@SundstromHenrySperling:2022]. Overall, fewer organizations work on contentious causes, instead focusing on safer issues like health and education. These changes in the civil society legal environment have influenced official donor responses. While multilateral donors do not reduce aid in response to anti-NGO laws, these laws are associated with a 32% decline in bilateral aid inflows [@DupuyPrakash:2018]. Further, advocacy-oriented donors---i.e., those that fund democracy and civil society promotion activities---reduced their spending by 74%, but did not cut their spending on development projects such as education, health, water, and sanitation [@RightSpringmanWibbels:2022].

@@

It is possible to study the effects of both organizational and structural heuristics on donor preferences individually, but disentangling the interaction of these heuristics adds complex dimensionality and makes more standard experimental work costly and statistically fraught. To address this, we use a conjoint experiment to measure the causal effect of both organizational and structural heuristics on individuals' preferences for donating to international NGOs. [@HainmuellerHopkinsYamamoto:2014]. Conjoint experiments are increasingly common in nonprofit studies, particularly for studying individual donor behavior [@SilviaChildWitesman:2023; @HirschmannMoritzBlock:2022; @BachkeAlfnesWik:2014].

A conjoint research design also allows us to estimate treatment effects even if respondents do not see every combination of organization attributes, and correspondingly provides a substantial increase in statistical power. In standard factorial experimental designs, participants would ordinarily need to be shown all possible combinations of experimental attributes. In our experiment, we presented respondents with six possible treatments with randomized attributes for each treatment, yielding 576 possible unique combinations of features. Even with respondents answering twelve iterations of the experiment, not every combination was seen. However, as long as all the possible organizational attributes are well randomized and there are no systematic biases toward specific options (i.e., more respondents select the first option because it is the first) or toward earlier iterations of attribute choices (i.e., respondents are more careful and attentive for the first hypothetical organization than the last), we can pool all observations together for specific attributes of interest while marginalizing across all other attributes [@KertzerRenshonYarhi-Milo:2021]. This allows us to (1) estimate the effect of each experimental treatment even if some unique combinations of attributes were unseen, and (2) use a much smaller sample size than would be required in a more traditional factorial design.

@@

[^prereg]: For clarity, we focus on our simplest, non-nested hypotheses, omitting many preregistered ones that overlap (e.g., the effect of branding and issue area and transparency and crackdown simultaneously). Following @WillrothAtherton:2024, we outline all the deviations from our preregistration protocol in the appendix.

@@

We fielded our survey experiment through Centiment, which recruits representative samples of paid (and highly engaged) survey participants online. To see how varying NGO characteristics influence the decision to donate, our sample was representative of the population of people who are likely to donate to charity. We asked potential participants a screening question about their philanthropic behavior early in the survey---if a participant responded that they give once every few years or never, they were disqualified from the study. We also included an attention check question early in the survey and removed respondents who failed the question. Importantly, these screening questions were presented *prior to* the experimental manipulation to avoid post-treatment bias [@MontgomeryNyhanTorres:2018a]. Following screening, we received 1,016 viable responses. Tables A1, A2, and A3 provide a detailed breakdown of the individual characteristics of our sample. In general, respondents were well balanced across all pre-treatment characteristics, including gender, age, education, income, and attitudes toward charity. Additionally, Table A4 compares sample characteristics with nationally representative estimates from the 2019 Current Population Survey (CPS). Our sample matches national proportions of age, marital status, and education, and contains slightly more men and slightly higher income levels than the general population. Due to the initial screening question that limited participants to those with an interest in charitable activity, respondents are far more likely to have donated, volunteered, or voted in the past year than population CPS estimates. Our reported causal effects are therefore not representative of the entire population and instead are generalizable to people interested in charitable giving.

@@

[^real-names]: We used real organization names rather than fictionalized names to allow the experimental task to better mirror reality [@McDonald:2020] and to reduce respondent cognitive burden when reflecting on organization characteristics [@MajnemerMeibauer:2022].

After presenting respondents with a set of three randomly generated organizations, we asked them which of the three they would be willing to donate to, along with an option for no selection (see @tbl-example for an example). This choice indicated donation intention and reflected the option the respondent preferred most within that set. We then repeated the process eleven more times for a total of twelve randomized iterations of hypothetical combinations of attributes, resulting in 12,192 completed experimental tasks (12 iterations × 1,016 respondents). This iterative process across twelve experimental tasks is typical of conjoint analysis and allows us to estimate individual-level differences using a multilevel model and determine both aggregate respondent preferences and the causal effect of each experimental feature on those preferences.

@@

We present the posterior distributions of the marginal means and AMCEs for each of our experimental conditions in @fig-all-results and provide posterior medians, credible intervals, and other model diagnostics in Tables A7 and A9. Because AMCEs are relative statements (i.e., contrasts between one feature level and a reference level), we try to use logical reference levels: for binary treatments like transparency and accountability, we calculate the difference between false and true levels; for ordered treatments like host country relationship, we calculate the differences between different levels of crackdown compared to no crackdown. To avoid imposing an artificial order on other unordered treatment variables, we report both marginal means and AMCEs [@LeeperHoboltTilley:2020].
